In-silico molecular enrichment and clearance of the human intracranial space

### 2024-11-28 paper/main.tex
Push some rough stuff for the introduction. WIP.

Molecular transport in perivascular spaces (PVSs) is established as a key pathway for human brain clearance and delivery. Previous studies indicate that molecules move rapidly in the subarachnoid space (SAS) and in PVSs surrounding pial arteries. Here, we aim to model and study this transport in a full pial perivascular and/or vascular network embedded in the SAS.

Motivation: as indicated by~\cite{iliff2012paravascular}, referring to
Cserr, Phys Rev., 1971, bulk convective flow may be more important for
larger species (larger distances) and larger molecules.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
\mer{Update: ~2800 words and 7 subsections - around 300-450 words each?}
\begin{enumerate}
\item
  ``A digital twin'' for molecular enrichment and clearance in human intracranial spaces
\item
  CSF flow, the role of production versus no/cardiac/respiratory
  dispersive effects on tracer trajectories, flow bifurcation. With and without tentorium cerebellum.
\item
  ``The role'' of PVS flow speed
\item
  ``The role'' of PVS sheaths and permeability - does functional separation entail structural separation?
\item
  ``The role'' of PVS size (motivated by pathologies)
\item
  Clearance from the CSF spaces - with and without low pressure veins, but keeping the intrathecal source of solute. 
\item
  Clearance from the parenchyma: $c_0 = 1.0$. Different substances?
\end{enumerate}

@@

@@

\newpage
\section*{Acknowledgments (not compulsory)}

>>>>

\section*{Introduction (rough ideas)}

% What is the problem? (Big picture)
% - Big picture/importance sentence.
% - More in-depth info
% - "XXX is unknown".

Transport of molecules within the spaces of the central nervous system
(CSF spaces, parenchyma) is important. Both in the context of
intracranial drug delivery with potential for personalized treatment
of cancer, diagnostics and treatment of CSF-related disorders, and in
the context of understanding clearance in connection with
neurodegenerative disease and for future treatment and diagnostics of
dementia disorders. Understanding and predicting molecular enrichment
is important for treatment and diagnostics of (brain) cancer, for
diagnostics of CSF-related disorders like normal pressure
hydrocephalus, subarachnoid hemorrage. Moreover, altered transport and
clearance is believed to be key factor in neurodegenerative diseases
like Alzheimer's disease and Parkinson's disease, and in sleep. CSF
flow and PVS established to play a key role. Clinical studies have
revealed large individual variation and variation between
cohorts/disease groups, therefore important in particular in the
context of personalized medicine. However, we are far from
understanding (the mechanisms behind) these variations.

% What we already know about the problem
Our current understanding has primarily been established via
experimental medical research (cats, rats, mice, pigs). Big open
questions and controversy remain, lack holistic understanding of these
physiological process remains, hard to translate experimental studies
to humans. In the last decade, clinical studies. Clinical studies have
revealed large individual variation and variation between
cohorts/disease groups. However, we are far from understanding the
mechanisms behind these variations and integrating and translating
experimental observations to humans.  In-silico modelling is starting
to play an integral role, evaluating hypothesis and pinpointing
mechanisms, but most efforts have been focused on local aspects and
lacking a holistic model. Hard to quantify the overall contribution of
individual effects.

PVS flow is important as shown by experimental studies. This is being
corroborated by clinical studies in humans (Eide and Ringstad (2024),
and the new one). However, a number of open questions remain: to what
extent do the PVS constitute an ensheathed space, how fast is the flow
in human PVS, what is the effect of dilated PVS as seen in disease
cases. Also, open questions on transport characteristics of the CSF
spaces - what gives the individual variations in tracer enhancements?

What is known is that fluid motion (in the ventricular system and SAS,
in perivascular spaces, and possibly within the parenchyma) and
pulsatility play key
roles. Glymphatics~\cite{iliff2012paravascular}. Diffusion, PVS flow
and dispersion, CSF flow. Molecular transport in perivascular spaces
(PVSs) is established as a key pathway for human brain clearance and
delivery. Previous studies indicate that molecules move rapidly in the
subarachnoid space (SAS) and in PVSs surrounding pial arteries. CSF
production in the choroid plexus and possibly in the parenchyma. CSF
efflux pathways, meningeal lympahtics, nerve routes. See also to
Cserr, Phys Rev., 1971, bulk convective flow may be more important for
larger species (larger distances) and larger molecules. Contribution
both from net flow (convection) and pulsatility (mixing) both in the
CSF spaces and in PVSs. Some aspects are better established, such as;
while other factors are more disputed such as ... contribution of
vasomotion. Moreover, the role of local effects in the bigger picture
has remained hard to evaluate.

% What are our research questions and how do we answer them
Recent developments on 3D-1D modelling of transport in PVS and
tissue. ``Digital twin for molecular enrichment and clearance in human
intracranial spaces''. Can we model the interplay between
molecular transport in the intracranial compartments, CSF flow and PVS
flow and transport to holistically quantify the contributions in humans?

% Main findings and outlook
Here, we aim to model and study this transport in a full pial
perivascular and/or vascular network embedded in the SAS.

\iffalse
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
\mer{Update: ~2800 words and 7 subsections - around 300-450 words each?}
\begin{enumerate}
\item
  ``A digital twin'' for molecular enrichment and clearance in human intracranial spaces
\item
  CSF flow, the role of production versus no/cardiac/respiratory
  dispersive effects on tracer trajectories, flow bifurcation. With and without tentorium cerebellum.
\item
  ``The role'' of PVS flow speed
\item
  ``The role'' of PVS sheaths and permeability - does functional separation entail structural separation?
\item
  ``The role'' of PVS size (motivated by pathologies)
\item
  Clearance from the CSF spaces - with and without low pressure veins, but keeping the intrathecal source of solute. 
\item
  Clearance from the parenchyma: $c_0 = 1.0$. Different substances?
\end{enumerate}

@@

\fi

@@

\iffalse
\newpage
\section*{Acknowledgments (not compulsory)}
